A Methodology and Implementation of Automated Emissions Harmonization for Use in Integrated Assessment Models

### 2017-05-19 meth.tex
updating paper and meth
saving work, meth and results mostly done

IAM emission results are provided along a number of different dimensions,
including, temporal (normally half decade or decade), spatial (i.e., model
regions), gas species, and sectoral. Each individual temporal trajectory (i.e.,
unique combinations of regions, species, and sectors) must be harmonized to the
initial time period. Given a model trajectory, $m_{r, g, s}(t)$, historical
trajectory, $h_{r, g, s}(t)$, and initial time period $t_i$, a harmonized
trajectory can be calcluated.

@@

\subsection{Default Method Decision Tree}

@@

\begin{figure}
  \begin{center}
    \includegraphics[width=\textwidth]{decision_tree.png}
    \caption[]{
      \label{fig:decision_tree}
      The default method decision tree used in the \code{aneris} software
      library. Decision-making threshold are described below, and the methods
      applied are described in Section \ref{sec:impl-meths}.}
  \end{center}
\end{figure}

A number of characteristics impact the decision of which default method to
select based on the effect of the characteristic on the potential harmonized
trajectory. For example, it is possible for models to report zero values in the
harmonization year in situations in which technologies are introduced in future
time periods in regions or for sectors which produce an emissions species that
is absent in the initial modeling period. In such cases, an offset method is
required as a ratio method would mask future emissions and erroneously harmonize
the trajectory. In most cases, however, models do report values in the
harmonization year. When model and historical values are relatively close, a
convergence method is provided in order to be as representative as possible to
the underlying unharmonized model results. The final model-based characteristic
of import is whether the model reports negative emissions. Such a case is
possible for gas species which can be extracted from the enviornment, as is the
case for CO2 in future scenarios with high mitigation strategies. If a model
provides a trajectory that transitions from positive to negative emissions, then
a convergence method is used in order to guarantee capture of this
transition. Otherwise, it is possible, given the relative values between history
and model in the harmonization year, for a negative trajectory to be
inappropriatley harmonized to positive, but decreasing, trajectory.

Variability of the historical trajectory is also an important characteristic
when considering the choice of harmonization method. Such trajectories are
normally indicative of emissions from land use sectors. Land use sectors are
modeled with varying fidelity by IAMs, but in general proxies are used for
emissions trajectories. Take for example the emissions of black carbon (BC) from
grassland burning in Africa \TODO{check that this is a good sector}: for most
models, the use of grass and paturelands is modeled, and the emissions of
species from the burning of these lands is determined by applying a emissions
factor to a proxy model of burning of these lands; in general, the more
grassland usage occurs, the higher the emissions from burning of
grasslands. Therefore, consistency in harmonization method is important due to
the approaches taken in modeling these sectors by IAMs. In order to reliably
detect land-use-like emissions, a measure of the covariance of the historical
trajectory is calculated as shown in Equation \ref{eqs:cov} and tested against a
threshold. To determine this threshold, an analysis of the recent CEDS
\TODO{cite ceds paper} historical data has been performed. Figure \ref{fig:cov}
shows the distribution of land-use covariances and non-land-use covariances as
determined for historical data aggregated to the model regions of 5 different
flagship IAMs \TODO{cite all 5 IAMs}. A threshold value of 20 has been chosen
based on these observations. Importantly, tails of the LUC and non-LUC overlap,
thus there are both false positives (~10\% of non-LUC trajectories) and false
negatives (~7\% of LUC trajectories). However, as any regional definition is
model depedent and thus any regional aggregation is possible an automated
detection mechanism is necessary.

@@

Finally, consideration is taken with respect to the relative difference between
the historic and model values in the harmonization year. In order to investigate
the possible values that these relative differences can take, the flagship model
values used in the SSP \TODO{CITE} and (ongoing) CMIP6 intercomparison exercises
are used. A distribution of these differences for all models in the study is
presented in Figure \ref{fig:dh}. Given the available data, a threshold value of
50\% was chosen to be used as a default in \code{aneris}.

@@

\begin{table}[]
\centering
\caption{Metadata provided by the \code{aneris} harmonization routine. This metadata is provided for every combination of region, sector, and emissions species.}
\label{my-tab:metadata}
\begin{tabular}{|p{2cm}|p{8cm}|}
\hline
Column       & Description                                                                  \\
\hline
\hline
method       & The harmonization method used.                                               \\
\hline
default      & The default harmonization method as determined by the default decision tree. \\
\hline
override     & The method provided as an override (if any).                                 \\
\hline
offset       & The offset value between history and model in the harmonization year.        \\
\hline
ratio        & The ratio value between history and model in the harmonization year.         \\
\hline
cov          & The covariance value of the historical trajectory.                           \\
\hline
unharmonized & The unharmonized value in the harmonization year.                            \\
\hline
history      & This historical value in the harmonization year.                             \\
\hline
harmonized   & The resulting harmonized value in the harmonization year.\\
\hline
\end{tabular}
\end{table}

>>>>

IAM emission results are provided along a number of different dimensions,
including, temporal (normally half decade or decade), spatial (i.e., model
regions), gas species, and sectoral. Each individual temporal trajectory (i.e.,
unique combinations of regions, species, and sectors) must be harmonized to the
initial time period. Given a model trajectory, $m_{r, g, s}(t)$, historical
trajectory, $h_{r, g, s}(t)$, and initial time period $t_i$, a harmonized
trajectory can be calculated.

@@

\subsection{Default Method Decision Tree}\label{sec:tree}

@@

\begin{figure}
  \begin{center}
    \includegraphics[width=\textwidth]{decision_tree.png}
    \caption[]{
      \label{fig:decision_tree}
      The default method decision tree used in the \code{aneris} software
      library. Decision-making threshold are described below.}
  \end{center}
\end{figure}

A number of characteristics impact the decision of which default method to
select based on the effect of the characteristic on the potential harmonized
trajectory. For example, it is possible for models to report zero values in the
harmonization year in situations in which technologies are introduced in future
time periods in regions or for sectors which produce an emissions species that
is absent in the initial modeling period. In such cases, an offset method is
required as a ratio method would mask future emissions and erroneously harmonize
the trajectory. In most cases, however, models do report values in the
harmonization year. When model and historical values are relatively close, a
convergence method is provided in order to be as representative as possible to
the underlying unharmonized model results. The final model-based characteristic
of import is whether the model reports negative emissions. Such a case is
possible for gas species which can be extracted from the environment, as is the
case for CO2 in future scenarios with high mitigation strategies. If a model
provides a trajectory that transitions from positive to negative emissions, then
a convergence method is used in order to guarantee capture of this
transition. Otherwise, it is possible, given the relative values between history
and model in the harmonization year, for a negative trajectory to be
inappropriately harmonized to positive, but decreasing, trajectory.

Variability of the historical trajectory is also an important characteristic
when considering the choice of harmonization method. Such trajectories are
normally indicative of emissions from land use sectors. Land use sectors are
modeled with varying fidelity by IAMs, but in general proxies are used for
emissions trajectories. Take for example the emissions of black carbon (BC) from
grassland burning in Africa \TODO{check that this is a good sector}: for most
models, the use of grass and paturelands is modeled, and the emissions of
species from the burning of these lands is determined by applying a emissions
factor to a proxy model of burning of these lands; in general, the more
grassland usage occurs, the higher the emissions from burning of
grasslands. Therefore, consistency in harmonization method is important due to
the approaches taken in modeling these sectors by IAMs. In order to reliably
detect land-use-like emissions, a measure of the co-variance of the historical
trajectory is calculated as shown in Equation \ref{eqs:cov} and tested against a
threshold. To determine this threshold, an analysis of the recent CEDS
\TODO{cite ceds paper} historical data has been performed. Figure \ref{fig:cov}
shows the distribution of land-use co-variances and non-land-use co-variances as
determined for historical data aggregated to the model regions of 5 different
flagship IAMs \TODO{cite all 5 IAMs}. A threshold value of 20 has been chosen
based on these observations. Importantly, tails of the LUC and non-LUC overlap,
thus there are both false positives (~10\% of non-LUC trajectories) and false
negatives (~7\% of LUC trajectories). However, as any regional definition is
model dependent and thus any regional aggregation is possible an automated
detection mechanism is necessary.

@@

Finally, consideration is taken with respect to the relative difference between
the historic and model values in the harmonization year. In order to investigate
the possible values that these relative differences can take, the flagship model
values used in the SSP \TODO{CITE} and (ongoing) CMIP6 inter-comparison exercises
are used. A distribution of these differences for all models in the study is
presented in Figure \ref{fig:dh}. Given the available data, a threshold value of
50\% was chosen to be used as a default in \code{aneris}.

@@

\begin{table}[]
\centering
\caption{Metadata provided by the \code{aneris} harmonization routine. This metadata is provided for every combination of region, sector, and emissions species.}
\label{tab:metadata}
\begin{tabular}{|p{2cm}|p{8cm}|}
\hline
Column       & Description                                                                  \\
\hline
\hline
method       & The harmonization method used.                                               \\
\hline
default      & The default harmonization method as determined by the default decision tree. \\
\hline
override     & The method provided as an override (if any).                                 \\
\hline
offset       & The offset value between history and model in the harmonization year.        \\
\hline
ratio        & The ratio value between history and model in the harmonization year.         \\
\hline
cov          & The covariance value of the historical trajectory.                           \\
\hline
unharmonized & The unharmonized value in the harmonization year.                            \\
\hline
history      & This historical value in the harmonization year.                             \\
\hline
harmonized   & The resulting harmonized value in the harmonization year.\\
\hline
\end{tabular}
\end{table}

### 2017-05-19 results.tex
result figures generated
saving work, meth and results mostly done

In order to test the \code{aneris} code base and harmonization procedure,
results from the IAM MESSAGE-GLOBIOM \TODO{CITE} was processed using the
previously described combined historical dataset. The SSP2-RCP4.5 scenario was
chosen a test cases as it incorporates moderate to heavy mitigation strategies,
and therefore produces negative CO2 emissions for some regions and sectors. In
general, SSP2 scenarios describe a ``middle of the road'' world within the SSP
framework context. The 4.5 designation implies a scenario that results in
approximately a 4.5 $\frac{\text{W}}{\text{m}^2}$ warming effect by the end of
the century, due to both warming from Greenhouse Gases (GHGs) and flourinated
gases as well as cooling from aerosols.

\subsection{The Use of Harmonization Overrides}

\subsection{Disccusion \& Future Work}

>>>>

In order to test the \code{aneris} harmonization procedure, results from the IAM
MESSAGE-GLOBIOM \TODO{CITE} was processed. Two scenarios from the Shared Socioeconomic
Pathways scenario library \TODO{CITE} are investigated. The reference SSP2, or
``middle of the road'' is chosen first because MESSAGE-GLOBIOM is the marker
scenario for this SSP. Notably, emissions from multiple sectors tend to increase
in SSP2, thus testing the harmonization mechanism with a increasing emission
pathway. Additionally, the SSP-45 scenario is chosen, where the ``45''
designation refers to a scenario with end-of-century emissions resulting in
approximately a 4.5 $\frac{\text{W}}{\text{m}^2}$ warming effect by the end of
the century, due to both warming from Greenhouse Gases (GHGs) and fluoridated
gases as well as cooling from aerosols. In this scenario, mitigation
technologies and policies are enacted causing a general reduction in pollutants
and GHGs, including (eventual) negative CO2 emissions in some regions and
sectors due to carbon capture and sequestration and afforestation. This scenario
explores the harmonization mechanism's response to generally decreasing emission
trajectories.

MESSAGE-GLOBIOM includes a representation of 11 distinct regions which can be
mapped directly to the 5-region definition used in the RCPs. Historical data is
taken from CEDS and LUC \TODO{get correct name}, which comprise 10 separate
pollutant and GHG species and 13 sectors. Therefore, a total of 1430 distinct
trajectories were harmonized.

The effect of harmonization overrides is analyzed by harmonizing each scenario
first without overrides and then with overrides. Of the 1430 trajectories,
approximately 2\% required the use of harmonization overrides after an initial
investigation; thus, 98\% of all trajectories were satisfactorily determined
using the default methods as chosen by the harmonization algorithm. Figure
\ref{fig:nox} provides one such example of a emissions species and sector in
which all regions were satisfactorily harmonized with the default methods. As
can be seen, results with and without overrides are identical.

\begin{figure}
  \begin{center}
    \includegraphics[width=\textwidth]{results_NOx_Energy_Sector.pdf}
    \caption[]{
      \label{fig:nox}
      NOx Energy Sector harmonized (solid lines) and unharmonized (dashed lines)
      trajectories for SSP2 and SSP2-45 are presented. SSP-45 is denoted with
      markers. Panel \textbf{a} shows 5-Region trajectories for scenarios
      \textit{without} any overrides. Panel \textbf{b} shows 5-Region
      trajectories for scenarios \textit{with} overrides. Panels \textbf{c} and
      \textbf{d} show global trajectories without and with overrides,
      respectively.  }
  \end{center}
\end{figure}

The harmonization of emissions pathways is performed in order to accurately
represent an updated historical emissions trajectory while also maintaining
consistency with the original, unharmonized pathway. When the default methods as
provided by the harmonization procedure distort or otherwise sufficiently
misrepresent the underlying unharmonized results, an override method is required
to be provided for the trajectory of the region, sector, and species in
question. Based on the 2\% of total trajectories required to be overridden, two
classifications were observed: regional trajectories whose \textit{magnitude}
was overly distorted and regional trajectories whose \textit{shape} was overly
distorted.

Figure \ref{fig:co} presents a situation in which the magnitude of a trajectory
is distorted. A large discrepancy (~300\% relative difference) is observed in
the harmonization year for carbon monoxide (CO) emissions in the industrial
sector specifically for the South Asia (SAS) MESSAGE-GLOBIOM region, which
comprises most of the Asian subcontinent. The default method chosen
(\code{constant_ratio} maintains model trends for the region, overall model
results are distorted. By applying a \code{constant_offset} override, the
regional trend and shape is maintained. With the new harmonization method for
the SAS region, the global trajectory for industrial CO also corresponds more
closely with the unharmonized trajectory.

\begin{figure}
  \begin{center}
    \includegraphics[width=\textwidth]{results_CO_Industrial_Sector.pdf}
    \caption[]{
      \label{fig:co}
      CO Industrial Sector harmonized and unharmonized emissions are presented
      for SSP2 and SSP2-45 scenarios. Scenarios as denoted identically to Figure
      \ref{fig:nox}. Panels \textbf{a} and \textbf{b} show harmonized and
      overridden-harmonized (respectively) regional trajectories for the 3
      MESSAGE-GLOBIOM regions that comprise the R5ASIA region: Centrally Planned
      Asia (CPA), Other Pacific Asia (PAS), and South Asia (SAS). Notably, the
      SAS regional trajectory displays a distorted trajectory due to the
      harmonization-year difference between history and model results in both
      scenarios. The distortion is large enough to affect global results, as
      shown in Panels \textbf{c} and \textbf{d}.  
}
  \end{center}
\end{figure}

In certain circumstances, the application of the default harmonization methods
can affect not only the magnitude but also the shape of regional
trajectories. Figure \ref{fig:nh3} shows emissions trajectories for ammonia
(NH3) resulting from the agriculture sector in Asia. Again, the SAS region shows
a large discrepancy in the harmonization year (> 150\% in this case). The
resulting trajectory harmonized with a \code{constant_ratio} method by default
provides a large increase after 2080 in the SSP2 reference scenario. Notably,
the SSP2-45 scenario is not affected to the same degree. While this distortion
affects the magnitude of the SAS trajectory, it largely affects the post-2080
shape of the global trajectory (see Figure \ref{fig:nh3}, panel \textbf{c}). By
using a \code{constant_offset} method as an override, this distortion is
addressed and more accurately reflects unharmonized results both in the SAS
region and global results for agricultural ammonia emissions.

\begin{figure}
  \begin{center}
    \includegraphics[width=\textwidth]{results_NH3_Agriculture.pdf}
    \caption[]{
      \label{fig:nh3}
      NH3 Agriculture harmonized and unharmonized emissions are presented for
      SSP2 and SSP2-45 scenarios. Scenarios and panel layouts are identical to
      Figure \ref{fig:co}. In this case, the SAS trajectory again shows not only
      a magnitude distortion, but also a shape distortion at the tail of the
      trajectory. Override methods have been applied to correct the distortion.
    }
  \end{center}
\end{figure}

\subsection{Discussion \& Future Work}
